---
title: "Route Term Gaps, Export EML, and Plan Catalog Publication"
teaching: 80
exercises: 55
---

:::::::::::::::::::::::::::::::::::::: questions

- What should I do when no existing term fits?
- What information makes a new-term request useful?
- How do review-state and publication-state validation differ?
- How do reviewed SDP fields become a valid EML file?
- How can I preview or perform a guarded KNB catalog upload?
- How should I create a later package version without discarding reviewed metadata?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Turn unresolved mappings into a term-gap plan.
- Route candidate terms to shared, DFO-specific, or local/profile governance.
- Run the strict SDP validation gate before export.
- Export the reviewed SDP as validated EML 2.2.
- Preview an exact KNB/DataONE deposit and distinguish a dry run from a persistent live upload.
- Write an intentional new package version from reviewed metadata.

::::::::::::::::::::::::::::::::::::::::::::::::

```{r metasalmon-available, include = FALSE}
# This chunk runs when the lesson is built. It proves the workshop's
# metasalmon dependency loads, and it is what keeps
# {metasalmon} registered as a lesson dependency for sandpaper's renv
# machinery. The workflow chunks below are display-only (eval = FALSE): they
# need a fully reviewed package, a completed EML sidecar, and - for the live
# call - credentials, none of which exist at build time.
library(metasalmon)
```

## A gap is not a failure

When no candidate term fits, do not force the closest match. Record the gap and decide where it belongs.

The package should keep enough evidence that maintainers can understand the request without a long follow-up interview.

## Build a term-gap plan in your software lane

After editing the metadata files, reload and validate the current package state.

::::::::::::::::::::::::::::::::::::: group-tab

### R

```r
pkg <- read_salmon_datapackage(pkg_path)
validate_salmon_datapackage(pkg_path, require_iris = FALSE)
```

Then rerun suggestions only for unresolved pieces.

```r
reviewed_dict <- suggest_semantics(
  df = pkg$resources,
  dict = pkg$dictionary,
  codes = pkg$codes,
  table_meta = pkg$tables,
  dataset_meta = pkg$dataset
)

gaps <- detect_semantic_term_gaps(reviewed_dict)

if (nrow(gaps) == 0) {
  message("No unresolved term gaps; there is no request plan to render.")
  requests <- tibble::tibble()
  request_plan <- tibble::tibble()
} else {
  requests <- render_ontology_term_request(
    gaps,
    scope = "auto",
    ask = FALSE,
    profile_name = "local-program"
  )

  reviewed_scopes <- requests$request_scope
  # Review every value. For example, route a DFO-specific row explicitly:
  # reviewed_scopes[2] <- "gcdfo"

  requests <- render_ontology_term_request(
    gaps,
    scope = "auto",
    ask = FALSE,
    profile_name = "local-program",
    scope_overrides = reviewed_scopes
  ) |>
    dplyr::mutate(
      ontology_repo = dplyr::if_else(
        request_scope %in% c("smn", "gcdfo"),
        ontology_repo,
        NA_character_
      )
    )

  request_plan <- requests |>
    dplyr::select(
      dataset_id,
      table_id,
      target_scope,
      target_row_key,
      target_sdp_field,
      dictionary_role,
      search_query,
      top_non_smn_label,
      top_non_smn_source,
      gap_detection_basis,
      llm_new_term_namespace,
      llm_escalated_from,
      placement_recommendation,
      placement_confidence,
      request_scope,
      ontology_repo
    )

  request_plan
  readr::write_csv(request_plan, file.path(pkg_path, "term-request-plan.csv"))
}
```

### Python

```python
from pathlib import Path

import pandas as pd
from metasalmonpy import (
    detect_semantic_term_gaps,
    read_salmon_datapackage,
    render_ontology_term_request,
    suggest_semantics,
    validate_salmon_datapackage,
)

pkg = read_salmon_datapackage(pkg_path)
validate_salmon_datapackage(pkg_path, require_iris=False)

reviewed_dict = suggest_semantics(
    df=pkg["resources"],
    dict_df=pkg["dictionary"],
    codes=pkg["codes"],
    table_meta=pkg["tables"],
    dataset_meta=pkg["dataset"],
)

gaps = detect_semantic_term_gaps(reviewed_dict)

if gaps.empty:
    print("No unresolved term gaps; there is no request plan to render.")
    requests = pd.DataFrame()
    request_plan = pd.DataFrame()
else:
    requests = render_ontology_term_request(
        gaps,
        scope="auto",
        ask=False,
        profile_name="local-program",
    )

    reviewed_scopes = requests["request_scope"].tolist()
    # Review every value. For example, route a DFO-specific row explicitly:
    # reviewed_scopes[1] = "gcdfo"

    requests = render_ontology_term_request(
        gaps,
        scope="auto",
        ask=False,
        profile_name="local-program",
        scope_overrides=reviewed_scopes,
    )

    non_repository_rows = ~requests["request_scope"].isin(
        ["smn", "gcdfo"]
    )
    requests.loc[non_repository_rows, "ontology_repo"] = pd.NA

    request_columns = [
        "dataset_id",
        "table_id",
        "target_scope",
        "target_row_key",
        "target_sdp_field",
        "dictionary_role",
        "search_query",
        "top_non_smn_label",
        "top_non_smn_source",
        "gap_detection_basis",
        "llm_new_term_namespace",
        "llm_escalated_from",
        "placement_recommendation",
        "placement_confidence",
        "request_scope",
        "ontology_repo",
    ]

    request_plan = requests.loc[:, request_columns]
    print(request_plan)
    request_plan.to_csv(
        Path(pkg_path) / "term-request-plan.csv",
        index=False,
    )
```

### Spreadsheet

Open `semantic_suggestions.csv` and the canonical metadata CSVs. For each unresolved field or code, record the dataset, table, field, search wording, closest candidate, why it does not fit, supporting source, and proposed route (`smn`, `gcdfo`, `profile`, or defer). Keep this as a reviewable term-gap plan; spreadsheet review does not submit a request.

::::::::::::::::::::::::::::::::::::::::::::::::

`detect_semantic_term_gaps()` combines deterministic candidate gaps with final LLM `request_new_term` decisions when LLM assessment was enabled. `gap_detection_basis` records which evidence created the gap, and `llm_escalated_from` preserves an unresolved `reject_shortlist` escalation. The deterministic `suggest_semantics()` call shown above leaves those LLM fields blank.

`render_ontology_term_request()` converts the gaps into reviewable draft payloads and proposed routes. It never submits an issue. The output is a working list, not an automatic governance decision.

To preview the SMN and GCDFO issue bodies without posting anything:

::::::::::::::::::::::::::::::::::::: group-tab

### R

```r
issue_preview <- if (nrow(requests) == 0) {
  tibble::tibble()
} else {
  requests |>
    dplyr::filter(request_scope %in% c("smn", "gcdfo")) |>
    submit_term_request_issues(dry_run = TRUE)
}

issue_preview
```

### Python

```python
from metasalmonpy import submit_term_request_issues

if requests.empty:
    issue_preview = pd.DataFrame()
else:
    routable = requests.loc[
        requests["request_scope"].isin(["smn", "gcdfo"])
    ]
    issue_preview = (
        submit_term_request_issues(routable, dry_run=True)
        if not routable.empty
        else pd.DataFrame()
    )

print(issue_preview)
```

### Spreadsheet

Use the term-gap plan as the draft review surface. A maintainer can turn an approved row into the relevant repository's request format later; nothing in the spreadsheet posts an issue or mints a term.

::::::::::::::::::::::::::::::::::::::::::::::::

## Routing decision

Use this plain-language triage:

| Route | Use when |
| --- | --- |
| `smn` | reusable salmon science concept, stable across organizations |
| `gcdfo` | DFO policy, operations, data stewardship, or DFO-specific concept |
| `profile` | local code list, project method bin, status category, formal local structure, or uncertain reuse |
| `skip` or defer | description is enough for now, or evidence is too weak |

Treat any LLM namespace suggestion as evidence, not authority. Review `request_scope` row by row before filing or publishing anything.

## Useful term-request ingredients

A useful request includes:

- proposed label;
- one or two sentence definition;
- source of the definition;
- example values or rows;
- where the term might sit in a hierarchy or concept scheme;
- related terms that were close but not exact;
- whether the term is shared, DFO-specific, or local/profile;
- any mapping strength you believe is appropriate.

## Final validation is later

Strict validation is the final gate. Use the language lane you selected earlier.

::::::::::::::::::::::::::::::::::::: group-tab

### R

```r
validate_salmon_datapackage(pkg_path, require_iris = TRUE)
```

### Python

```python
validate_salmon_datapackage(pkg_path, require_iris=True)
```

### Spreadsheet

Spreadsheet review can resolve metadata content, but it cannot execute the strict validator. Run this final gate against the same package folder in R or Python before EML export or publication.

::::::::::::::::::::::::::::::::::::::::::::::::

Run it only when:

- review markers such as `REVIEW:` have been resolved;
- required metadata placeholders are gone;
- canonical metadata headers exactly match the schema fields, in schema order, with no missing or extra columns;
- every `tables.csv$file_name` is a safe relative path to an existing data file;
- `column_dictionary.csv` rows and column names match the headers of the referenced data files;
- every observed non-empty categorical value has exactly one matching `codes.csv` row for its dataset/table/column/value key;
- measurement rows have final `term_iri`, `property_iri`, `entity_iri`, and `unit_iri`;
- `datapackage.json` exists, declares the SDP profile and resources, and agrees with the canonical metadata CSVs and referenced data files;
- term requests or local/profile decisions are documented.

Deferred measurement gaps are valid in draft or review state. They will not pass strict publication validation until the required measurement semantic fields are resolved.

## Map the reviewed SDP into EML

`metasalmon` and `metasalmonpy` do not simply rename SDP fields. They build a new EML document and validate the result:

| Reviewed SDP source | EML output |
| --- | --- |
| Dataset ID, title, description, keywords, temporal coverage | Alternate identifier, title, abstract, keyword set, and temporal coverage |
| Each table row and referenced data CSV | One EML `dataTable` with an identified physical object |
| Column name, label, description, and `value_type` | Attribute name, label, definition, and storage type |
| `required = TRUE` and table primary keys | Not-null and primary-key constraints |
| Categorical `codes.csv` rows | Enumerated-domain code definitions |
| Reviewed measurement term and unit IRIs | Conservative semantic annotations |
| Reviewed EML sidecar | Parties, rights, methods, measurement scales, missing values, geographic/taxonomic coverage, provenance, and access intent |

The canonical SDP does not contain every fact EML requires. For example, a free-text `creator` cell cannot safely be guessed into structured people, organizations, ORCIDs, and metadata-provider roles. Measurement scale and the meanings of missing-value tokens also need explicit review.

## Export validated EML

The code blocks from here to the end of this section operate on `pkg_path`, the reviewed package you have been building through the workshop; they are shown for you to run in your own session and are not executed when the lesson website is built.

Create `metadata/eml-mapping.yml` from metasalmon's [canonical mapping template][metasalmon-eml-mapping]. R users can copy the template from their installed package with the code below; Python and spreadsheet users can download the linked file and save it at the same package-relative path.

```{r, eval = FALSE, purl = FALSE}
eml_mapping_path <- file.path(
  pkg_path,
  "metadata",
  "eml-mapping.yml"
)

if (!file.exists(eml_mapping_path)) {
  file.copy(
    system.file(
      "extdata",
      "eml-mapping-template.yml",
      package = "metasalmon"
    ),
    eml_mapping_path
  )
}
```

Stop and review the copied YAML. Replace every example value. The sidecar must bind the final semantic vocabulary and reviewed-selection ledger with their SHA-256 checksums; it must also record reviewed parties, rights authorization, source provenance, methods, and one measurement-scale/domain entry for every column. Set `status: final` only after that review.

The detailed, current checklist is in the [metasalmon post-review and publication workflow][metasalmon-eml-workflow]. Do not invent dummy parties, checksums, rights evidence, measurement scales, or missing-value meanings merely to make validation pass.

When the SDP, its semantic-review evidence, and the sidecar are final, use your software lane:

::::::::::::::::::::::::::::::::::::: group-tab

### R

```{r, eval = FALSE, purl = FALSE}
eml_result <- write_eml_from_sdp(pkg_path)

eml_result$path
eml_result$eml_version
eml_result$validation
```

### Python

```python
from metasalmonpy import write_eml_from_sdp

eml_result = write_eml_from_sdp(pkg_path)

print(eml_result["path"])
print(eml_result["eml_version"])
print(eml_result["validation"])
```

### Spreadsheet

Spreadsheet software can be used to complete and review the SDP metadata and `eml-mapping.yml`, but it does not run the EML schema validator. When you are ready for export, open the same package folder in either code lane and run its exporter; the reviewed package remains the source state.

::::::::::::::::::::::::::::::::::::::::::::::::

The default output is `metadata/eml.xml`. An identical existing file is an idempotent success. A different existing file is protected unless you deliberately pass `overwrite = TRUE` / `overwrite=True` after reviewing why the bytes changed.

## Preview the KNB/DataONE catalog deposit

Use a dry run first. It creates local EML, OAI-ORE, and manifest artifacts, but it does not read credentials or make a network request:

::::::::::::::::::::::::::::::::::::: group-tab

### R

```{r, eval = FALSE, purl = FALSE}
knb_plan <- publish_sdp_to_knb(
  pkg_path,
  public = FALSE,
  dry_run = TRUE,
  representation = "expanded",
  overwrite = FALSE,
  knb_environment = "test"
)

knb_plan$status
knb_plan$manifest_path
```

### Python

```python
from metasalmonpy import publish_sdp_to_knb

knb_plan = publish_sdp_to_knb(
    pkg_path,
    public=False,
    dry_run=True,
    representation="expanded",
    overwrite=False,
    knb_environment="test",
)

print(knb_plan["status"])
print(knb_plan["manifest_path"])
```

### Spreadsheet

Spreadsheet software cannot generate the cryptographic manifest. Review an instructor-provided dry-run manifest alongside the package: confirm the intended files, identifiers, checksums, selected environment, and access decision. Generate a new manifest in R or Python before any real deposit.

::::::::::::::::::::::::::::::::::::::::::::::::

Review `publication/test/knb-manifest.json`. The expanded plan lists the original data resources, allowlisted canonical SDP artifacts, validated EML science metadata, and the OAI-ORE resource map with exact identifiers and checksums. It does not scan and upload arbitrary files from the package folder. The test environment is a rehearsal: it is non-durable, cannot be promoted into production, is unsuitable for sensitive data, and cannot receive a DOI.

`overwrite = TRUE` / `overwrite=True` can rebuild conflicting local artifacts left by an **unpublished dry run** after you correct an input. It cannot overwrite anything that reached DataONE; published PIDs are immutable and require a reviewed revision.

## Live upload is a separate authorized action

A live call mutates the selected KNB environment. A production call creates persistent KNB objects even when `public = FALSE`; private access is a review posture, not a server-side draft. Rehearse against the test node first, then create and review a separate production dry-run plan before authorizing production. The production call requires:

- the exact pre-existing production dry-run plan;
- a short-lived production DataONE JWT entered only into the current process;
- an authenticated subject matching the reviewed EML metadata provider;
- `confirm = TRUE`; and
- authority for the chosen access and redistribution decision.

The code below is intentionally display-only. It first creates the credential-free production plan that must be reviewed, then shows the separate live action. Here `overwrite = TRUE` / `overwrite=True` permits the planner to replace the unpublished standalone EML with the publication-aware EML that names the planned objects; it is not permission to overwrite anything already deposited.

::::::::::::::::::::::::::::::::::::: group-tab

### R

```{r, eval = FALSE, purl = FALSE}
production_plan <- publish_sdp_to_knb(
  pkg_path,
  public = FALSE,
  dry_run = TRUE,
  representation = "expanded",
  overwrite = TRUE,
  knb_environment = "production"
)

production_plan$manifest_path

previous_token <- getOption("dataone_token")

options(
  dataone_token = rstudioapi::askForPassword(
    "Short-lived DataONE JWT"
  )
)

tryCatch(
  publish_sdp_to_knb(
    pkg_path,
    public = FALSE,
    dry_run = FALSE,
    confirm = TRUE,
    representation = "expanded",
    knb_environment = "production"
  ),
  finally = options(dataone_token = previous_token)
)
```

### Python

```python
from getpass import getpass

from metasalmonpy import publish_sdp_to_knb
from metasalmonpy.knb_publication import set_dataone_token

production_plan = publish_sdp_to_knb(
    pkg_path,
    public=False,
    dry_run=True,
    representation="expanded",
    overwrite=True,
    knb_environment="production",
)

print(production_plan["manifest_path"])

set_dataone_token(getpass("Short-lived production DataONE JWT: "))

try:
    publication_result = publish_sdp_to_knb(
        pkg_path,
        public=False,
        dry_run=False,
        confirm=True,
        representation="expanded",
        knb_environment="production",
    )
finally:
    set_dataone_token(None)
```

### Spreadsheet

Do not perform a live upload from a spreadsheet. Use the manifest as the approval surface, then run the exact reviewed plan in R or Python only when the data owner and publisher have authorized the production access and redistribution decision.

::::::::::::::::::::::::::::::::::::::::::::::::

Do not place the token in a script, `.Renviron`, YAML, manifest, command argument, or shell history. Do not switch to `public = TRUE` merely for convenience: public access is a separate redistribution decision and requests DataONE preservation replicas.

Publication is complete only when the intended catalog state is verified, not merely when KNB stores the objects. If the returned manifest says `published_pending_catalog`, preserve that exact status and recheck rather than creating replacement identifiers or claiming completion.

## Later example: make a new version after review

After people have edited package metadata over time, do not regenerate the same folder from raw data and hope the edits survive. Read the reviewed state, make deliberate changes in memory, and write a new versioned folder:

::::::::::::::::::::::::::::::::::::: group-tab

### R

```r
reviewed_pkg <- read_salmon_datapackage(pkg_path)

# Example of an intentional metadata update. Replace with the real reviewed text.
reviewed_pkg$dataset <- reviewed_pkg$dataset |>
  dplyr::mutate(
    description = paste(
      description,
      "This version adds the reviewed 2026 return-year records."
    )
  )

v2_path <- file.path("output", "my-salmon-sdp-v2")

write_salmon_datapackage(
  resources = reviewed_pkg$resources,
  dataset_meta = reviewed_pkg$dataset,
  table_meta = reviewed_pkg$tables,
  dict = reviewed_pkg$dictionary,
  codes = reviewed_pkg$codes,
  path = v2_path,
  overwrite = FALSE
)

validate_salmon_datapackage(
  v2_path,
  require_iris = TRUE
)
```

### Python

```python
from pathlib import Path

from metasalmonpy import (
    read_salmon_datapackage,
    validate_salmon_datapackage,
    write_salmon_datapackage,
)

reviewed_pkg = read_salmon_datapackage(pkg_path)

reviewed_dataset = reviewed_pkg["dataset"].copy()
reviewed_dataset.loc[:, "description"] = (
    reviewed_dataset["description"].astype(str)
    + " This version adds the reviewed 2026 return-year records."
)
reviewed_pkg["dataset"] = reviewed_dataset

v2_path = write_salmon_datapackage(
    resources=reviewed_pkg["resources"],
    dataset_meta=reviewed_pkg["dataset"],
    table_meta=reviewed_pkg["tables"],
    dict_df=reviewed_pkg["dictionary"],
    codes=reviewed_pkg["codes"],
    path=Path("output") / "my-salmon-sdp-v2",
    overwrite=False,
)

validate_salmon_datapackage(
    v2_path,
    require_iris=True,
)
```

### Spreadsheet

Preserve the reviewed package unchanged. Copy it to a clearly named new version folder, make the deliberate metadata or data changes in that copy, and document what changed. Before EML export or publication, run strict validation against the new folder in R or Python; do not treat a manual copy as validated merely because its files open successfully.

::::::::::::::::::::::::::::::::::::::::::::::::

If data columns changed, update the dictionary. If new categorical values appeared, update `codes.csv`. Rebuild and review version-specific semantic, EML, and publication artifacts; do not blindly copy old checksum-bound sidecars.

For a later KNB version, keep the earlier package and verified manifest unchanged, add a new reviewed `publication.revision_key` to the new version's EML sidecar, and plan against the preceding `revision_manifest`. This creates an immutable revision rather than overwriting published bytes.

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Draft a term-gap plan

Choose one unresolved field or code value.

Decide:

- route: `smn`, `gcdfo`, `profile`, or `skip`/defer;
- proposed label and definition;
- evidence source;
- why existing terms do not fit;
- next action.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Trace the publication workflow

Using a finalized instructor-provided package or your own package when it is ready:

1. identify three SDP fields and their EML destinations;
2. identify two facts that must come from `eml-mapping.yml` rather than inference;
3. run or inspect the result of `write_eml_from_sdp()`;
4. run or inspect a credential-free KNB dry-run manifest; and
5. state who would need to authorize a live upload and whether the intended access is private or public.

Do not run a live upload as a classroom experiment.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- Missing terms are governance work, not a reason to force bad mappings.
- High-quality requests carry definitions, sources, examples, and routing rationale.
- Rendering produces reviewable SMN, GCDFO, or profile request drafts; it does not submit them.
- Strict validation is a publication-readiness check, not the first review step.
- EML export is a validated transformation that also requires reviewed sidecar facts.
- A KNB dry run is local and reversible; a live call creates persistent production objects.
- Later package and catalog versions should be explicit immutable revisions, not overwrite attempts.

::::::::::::::::::::::::::::::::::::::::::::::::
