---
title: "Review Terms and Measurement Meanings"
teaching: 55
exercises: 40
---

:::::::::::::::::::::::::::::::::::::: questions

- Where does the record of a semantic decision live?
- How do we keep semantic review small enough to finish in the workshop?
- How do I decide whether a suggested IRI is good enough to keep?
- What does I-ADOPT clarify for measurement columns?
- What can this review *not* decide for me?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Build a review queue from the suggestions your package already carries, without running a new search.
- Read one slot's shortlist and decide it with a call you paste into `scripts/build_sdp.R`.
- Decompose one measurement once using the SDP's I-ADOPT fields.
- Write the decisions back with `apply_sdp_semantics()` and confirm the data files did not change.
- Name the decisions this review cannot reach, and say where they are still made.

::::::::::::::::::::::::::::::::::::::::::::::::

```{r metasalmon-available, include = FALSE}
# Runs when the lesson site is built. It proves the workshop's metasalmon
# dependency loads and keeps {metasalmon} registered for sandpaper's renv
# machinery, exactly as the same chunk does in Session 6.
#
# It deliberately does NOT assert that review_semantics() exists. The lesson's
# renv lockfile pins one released metasalmon, and a hard assertion here would
# turn "the pin is behind the release" into a failed website build. The note
# below is the softer signal instead: it goes to stderr rather than through
# message(), because include = FALSE discards captured messages and this one is
# meant for the build log.
#
# Retires when: the lockfile pins a metasalmon that exports review_semantics().
# Delete the whole `if` at that point; the library() call stays.
library(metasalmon)

if (!exists("review_semantics", where = asNamespace("metasalmon"), inherits = FALSE)) {
  cat(
    "NOTE: pinned metasalmon ", as.character(utils::packageVersion("metasalmon")),
    " does not export review_semantics(); session-4.Rmd teaches it. ",
    "Update renv/profiles/lesson-requirements/renv.lock.\n",
    sep = "", file = stderr()
  )
}
```

## The decision that used to leave no record

By the end of Session 3 your package was reproducible almost everywhere. `raw_data/` holds unchanged inputs, `scripts/build_sdp.R` turns them into a package, and rerunning the script rebuilds the same bytes.

Almost everywhere. Until now, the documented way to accept a semantic link was to open `metadata/column_dictionary.csv` in a spreadsheet, read `semantic_suggestions.csv` beside it, and copy an IRI across by hand. That worked. But the only record it left was the changed cell. Six months later the cell says *what* was decided and nothing at all about *why*, *by whom*, or *what else was on the shortlist*. It was the one unreproducible link in a chain that is otherwise byte-for-byte reproducible.

This chapter replaces that step. `review_semantics()` prints a shortlist and, under every candidate, the exact line of R that accepts it:

```output
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "variable", rank = 1)
```

You paste that line into `scripts/build_sdp.R`. **The paste is the audit trail.** There is no prompt, no menu, and no interactive review screen, and that is a design decision rather than an omission: anything you answer at a prompt vanishes when the session ends, which would leave the decision exactly as unreproducible as the spreadsheet it replaces.

A learner who understands why the spreadsheet was abandoned will not drift back to it. The spreadsheet lane still exists below, because a mixed workshop needs it — but it is now the lane that costs you the record, and this chapter says so plainly.

::::::::::::::::::::::::::::::::::::: prereq

## You need a metasalmon new enough to have the review flow

`review_semantics()` and its companions are recent. Check before you start, and reinstall from the [metasalmon repository][metasalmon] if the answer is `FALSE`:

```{r, eval = FALSE, purl = FALSE}
packageVersion("metasalmon")
exists("review_semantics", where = asNamespace("metasalmon"), inherits = FALSE)
```

The [changelog][metasalmon-changelog] names the release that introduced it.

::::::::::::::::::::::::::::::::::::::::::::::::

## Keep the workshop semantic review bounded

We will not try to map every column. Everyone will use the NuSEDS Fraser Coho escapement package from Session 2, even if you are also building a package from your own data. That gives the room one case to discuss and keeps the decisions small enough to finish.

We will decide exactly these linked slots:

| Decision | SDP destination | Shared Fraser Coho slot |
| --- | --- | --- |
| One complete measurement variable | `column_dictionary.csv$term_iri` | `NATURAL_SPAWNERS_TOTAL` |
| One I-ADOPT decomposition | `property_iri`, `entity_iri`, `constraint_iri` on that same row | abundance, of a population, natural-origin |
| One unit | `column_dictionary.csv$unit_iri` | a count |
| One statistical modifier | `column_dictionary.csv$statistical_modifier_iri` | a total |
| What each table row represents | `tables.csv$observation_unit_iri` | an escapement |
| One deliberate rejection | any slot | `AREA` |

Administrative IDs, file names, local notes, and the other categorical code lists can rely on clear descriptions for now and be mapped in a later review.

## Your package must already carry suggestions

`review_semantics()` **never searches and never contacts a network or an LLM.** It reads the suggestions your package already has. That is why Session 3 turned `seed_semantics = TRUE` on: without it there is nothing to review, and the function says so rather than quietly returning an empty queue.

```output
Error in review_semantics(pkg_path) : No semantic suggestions to review.
ℹ Run `suggest_semantics()`, or `create_sdp()` with `seed_semantics = TRUE`,
  first.
```

If the Session 2 package was built with `seed_semantics = FALSE`, rebuild it once with seeding on before continuing. Expect the search itself to take a few minutes; that cost is paid once, at build time, and never again during review.

## Build the review queue

::::::::::::::::::::::::::::::::::::: group-tab

### R

```{r, eval = FALSE, purl = FALSE}
library(metasalmon)

pkg_path <- file.path("output", "fraser-coho-example-sdp")

# One row per candidate; one queue entry per undecided slot. No search runs.
review <- review_semantics(pkg_path)

review
```

To work through one column at a time, filter the queue:

```{r, eval = FALSE, purl = FALSE}
review <- review_semantics(pkg_path, columns = "NATURAL_SPAWNERS_TOTAL")
```

### Python

`metasalmonpy` does not have this API yet. The R-native review flow lands in `metasalmon` first, and the Python equivalent is pending; there is nothing to demonstrate here that would still be true next month.

What a Python user can do today is read the same evidence the R queue is built from. `semantic_suggestions.csv` sits at the package root and is written by both implementations:

```python
from pathlib import Path

import pandas as pd

pkg_path = Path("output") / "fraser-coho-example-sdp"

suggestions = pd.read_csv(pkg_path / "semantic_suggestions.csv")

print(
    suggestions[
        [
            "column_name",
            "dictionary_role",
            "target_sdp_file",
            "target_sdp_field",
            "label",
            "iri",
            "source",
            "score",
        ]
    ]
)
```

Decide from that table, then edit the target field in the canonical metadata CSV. Record each decision in `scripts/build_sdp.py` as a comment or an explicit assignment, so the reasoning survives the way the pasted R call does. Until the Python review flow ships, that discipline is manual rather than enforced.

### Spreadsheet

Open `semantic_suggestions.csv` at the package root beside `metadata/column_dictionary.csv`. Filter to `NATURAL_SPAWNERS_TOTAL` and to the `tables.csv` row whose `target_sdp_field` is `observation_unit_iri`. Compare each `label`, `iri`, and `definition` with the source documentation and the data holder's explanation, then write the accepted IRI into the target field named by `target_sdp_file` and `target_sdp_field`.

Be clear about the cost. This lane records the *outcome* and not the *reasoning*: nothing in the saved file says which other candidates you saw, or why you preferred this one. Keep a written review log beside the package so the next reader is not guessing. Opening the CSV does not accept a suggestion and is not evidence that a link is correct.

::::::::::::::::::::::::::::::::::::::::::::::::

## Read one slot

Each queue entry is one block. This is the `NATURAL_SPAWNERS_TOTAL` variable slot — the first block you see after filtering with `columns =` — printed exactly as it appears:

```output
── escapement · NATURAL_SPAWNERS_TOTAL · variable ────────────────────────────
   field:   column_dictionary.csv · term_iri
   current: REVIEW: https://w3id.org/gcdfo/salmon#SpawnerAbundance

   [1] Spawner abundance   gcdfo   score 18.25
       The abundance (count) of adult salmon that reach spawning grounds in a
       given system and time period (often operationalized as
       escapement/spawning escapement).
       https://w3id.org/gcdfo/salmon#SpawnerAbundance
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "variable", rank = 1)

       review <- reject_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "variable")   # no candidate fits
```

Read it in this order:

- **the heading** — table, column, and semantic role;
- **`field:`** — the exact file and column this decision writes into, so you can check the result yourself;
- **`current:`** — what the field holds *now*;
- **the shortlist** — `[1]`, `[2]`, … in the order the ranked search produced, each with its label, source vocabulary, score, definition, and IRI; and
- **the calls** — one per candidate, plus one rejection.

`current:` has three states, and the difference matters.

| `current:` shows | What it means |
| --- | --- |
| `REVIEW: <iri>` | Seeding wrote a draft, because its unattended compatibility check passed. Still undecided. |
| `<blank>` | Seeding found a candidate but declined to draft it. Still undecided, and the queue shows it for exactly that reason. |
| `<unknown>` | The queue could not read the current value — the metadata row was missing or matched ambiguously. Read the CSV before deciding. |

Undecided is undecided. A `REVIEW:` prefix is a marker meaning *nobody has looked at this*, not a weak endorsement, and strict validation refuses to publish a package that still contains one.

The variable answers "what complete measurement is this column?" For `NATURAL_SPAWNERS_TOTAL` the candidate says spawner abundance, from `gcdfo`, with a definition that matches the NuSEDS field. Read the definition, not the label. Then accept it.

## Decide, in your script

`accept_suggestion()` and `reject_suggestion()` take a review and return a review. Nothing is written to disk until you say so, so a whole review is an ordinary re-runnable R script.

::::::::::::::::::::::::::::::::::::: group-tab

### R

```{r, eval = FALSE, purl = FALSE}
review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "variable", rank = 1)
```

Reject when nothing on the shortlist fits, and say why:

```{r, eval = FALSE, purl = FALSE}
review <- reject_suggestion(
  review,
  "AREA",
  "variable",
  reason = "DFO statistical area, not a body of water"
)
```

`reason` is carried on the review object and shown in the console, and it reaches the package only as the word `rejected` in `semantic_suggestions.csv` — the sentence itself is not stored there. So the reason survives in exactly one place: the line you paste into your script. Write it as if that were true, because it is.

Accept a term the search did not surface by naming it directly instead of a rank:

```{r, eval = FALSE, purl = FALSE}
review <- accept_suggestion(
  review,
  "NATURAL_SPAWNERS_TOTAL",
  "property",
  iri = "https://w3id.org/smn/Abundance"
)
```

### Python

Pending, as above. Apply the decision by editing the target field in the canonical metadata CSV named by `target_sdp_file` and `target_sdp_field`, and record the reasoning in `scripts/build_sdp.py`.

### Spreadsheet

Type the accepted IRI into the target field, with no `REVIEW:` prefix. Leave the field blank for a rejection and write the reason in your review log.

::::::::::::::::::::::::::::::::::::::::::::::::

Two things make the printed call safe to paste rather than merely suggestive. The argument set is computed by *resolving* it, so `table =` and `code_value =` appear only when they are needed to address one slot and never otherwise. And the object name is read from your own binding — bind the review to `sem_review` and the printed calls say `sem_review <- accept_suggestion(sem_review, …)`.

Mistakes are caught by name, not by silence:

```output
Error in accept_suggestion(sem_review, "FULL_CU", "variable", rank = 1) :
  No review slot matches that column and role.
✖ Asked for: FULL_CU · variable
ℹ FULL_CU_IN · variable
```

```output
Error in accept_suggestion(sem_review, "FULL_CU_IN", "variable", rank = 3) :
  No candidate with that `rank` in this slot.
ℹ Ranks available: 1.
ℹ To accept a term that is not shortlisted, pass `iri` instead.
```

Read `Ranks available: 1.` literally. In the seeded Fraser Coho package, `semantic_suggestions.csv` carries exactly one row per slot rather than a whole ranked list, so every shortlist is one candidate deep and `rank = 1` is the only rank there is. "Accept rank 1" is therefore a much weaker statement than "the best of five" — it is closer to "accept or reject". When rank 1 is wrong, `iri =` is the way out, not rank 2.

## Treat suggestions as drafts

A score is a retrieval score. It measures how well a query matched a label and a definition; it is not agreement, and it is not evidence about your column.

Here is the very first slot in the unfiltered Fraser Coho queue:

```output
── escapement · AREA · variable ──────────────────────────────────────────────
   field:   column_dictionary.csv · term_iri
   current: <blank>

   [1] water body   ols   score 9.35
       The term body of water most often refers to large accumulations of
       water, such as oceans, seas, and lakes, but it includes smaller pools
       of water such as ponds, wetlands, or more rarely, puddles. A body of
       water does not have to be still or contained; Rivers, streams, canals,
       and other geographical features where water moves from one place to
       another are also considered bodies of water.
       http://purl.obolibrary.org/obo/ENVO_00000063  https://www.ebi.ac.uk/ols4/ontologies/envo/classes/http%3A%2F%2Fpurl.obolibrary.org%2Fobo%2FENVO_00000063
       review <- accept_suggestion(review, "AREA", "variable", rank = 1)

       review <- reject_suggestion(review, "AREA", "variable")   # no candidate fits
```

`AREA` in NuSEDS is a DFO statistical management area — an administrative identifier. The top and only candidate is an ENVO class for a body of water. It scores 9.35 and it is wrong. Reject it.

`WATERSHED_CDE` is the sharper case, and it is the reason step 2 of the checklist below is not redundant with step 1. Its top candidate is labelled **watershed**, which matches the column name about as well as a label can. Read the definition and it turns out to be *"the separation between neighbouring drainage basins"* — the divide, not the basin. The column stores neither: it stores a code. The label matched and the concept did not, which is precisely the failure a label-only check cannot see.

Keep a suggested IRI only when:

1. the label matches what the column means;
2. the definition matches the dataset context;
3. the scope is right;
4. the unit is compatible with the values;
5. the mapping does not erase an important caveat.

If any check fails, reject it, supply the right IRI with `iri =`, or leave it for the term-gap workflow in Session 6.

::::::::::::::::::::::::::::::::::::: callout

## Where the second IRI on the line comes from

Candidates retrieved through OLS print their IRI followed by a browsable OLS4 URL, because the IRI itself does not resolve to a human-readable page. In a terminal that supports hyperlinks you get one clickable label instead. Candidates from `smn`, `gcdfo`, and NVS print one IRI, because that IRI is already the page.

::::::::::::::::::::::::::::::::::::::::::::::::

## Measurement semantics and I-ADOPT

For `column_role = measurement`, SDP combines a variable link, I-ADOPT-style meaning components, a unit link, and an optional statistical-modifier link.

| Field | Plain-language question |
| --- | --- |
| `term_iri` | What complete measurement variable is this? |
| `property_iri` | I-ADOPT property: what characteristic is measured — count, length, mass, temperature? |
| `entity_iri` | I-ADOPT entity/object of interest: what is being measured? |
| `constraint_iri` | Optional I-ADOPT constraint: what qualifier narrows the meaning? |
| `unit_iri` | What unit are the values in? |
| `statistical_modifier_iri` | Optional: is the column an aggregation — a mean, maximum, minimum, total, or peak — rather than a single observation? |

The queue shows one slot per field, so the decomposition is decided one question at a time.

```output
── escapement · NATURAL_SPAWNERS_TOTAL · property ────────────────────────────
   field:   column_dictionary.csv · property_iri
   current: REVIEW: https://w3id.org/gcdfo/salmon#SpawnerAbundance

   [1] Spawner abundance   gcdfo   score 18.95
       The abundance (count) of adult salmon that reach spawning grounds in a
       given system and time period (often operationalized as
       escapement/spawning escapement).
       https://w3id.org/gcdfo/salmon#SpawnerAbundance
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "property", rank = 1)
```

This one deserves a stop. Retrieval offered the *same term* for `term_iri` and `property_iri`, and it cannot be right in both places: "spawner abundance" is a whole measurement variable, whereas the I-ADOPT property is the bare characteristic being measured. Decomposing a variable into itself records nothing. The property here is abundance:

```{r, eval = FALSE, purl = FALSE}
review <- accept_suggestion(
  review,
  "NATURAL_SPAWNERS_TOTAL",
  "property",
  iri = "https://w3id.org/smn/Abundance"
)
```

This is the normal case for `iri =`, and it is worth naming as a general habit: when the property and the variable retrieve the same term, the property slot almost certainly needs a broader one.

The remaining three slots:

```output
── escapement · NATURAL_SPAWNERS_TOTAL · entity ──────────────────────────────
   field:   column_dictionary.csv · entity_iri
   current: REVIEW: https://w3id.org/smn/Population

   [1] Population   smn   score 15.55
       A group of organisms of the same species occupying a defined area that
       interbreed and share a gene pool. In salmon, populations are composed
       of one or more demes and form the biological basis for Conservation
       Units.
       https://w3id.org/smn/Population
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "entity", rank = 1)

── escapement · NATURAL_SPAWNERS_TOTAL · unit ────────────────────────────────
   field:   column_dictionary.csv · unit_iri
   current: REVIEW: http://qudt.org/vocab/unit/COUNT

   [1] Count   qudt   score 4.65
       http://qudt.org/vocab/unit/COUNT
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "unit", rank = 1)

── escapement · NATURAL_SPAWNERS_TOTAL · constraint ──────────────────────────
   field:   column_dictionary.csv · constraint_iri
   current: REVIEW: https://w3id.org/smn/NaturalOrigin

   [1] Natural-origin   smn   score 8.6
       Individuals that are born and reared in the wild (natural
       environment).
       https://w3id.org/smn/NaturalOrigin
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "constraint", rank = 1)
```

Note the unit candidate: score 4.65 and no definition at all. A low score with nothing to read is a prompt to check the vocabulary yourself, not a reason to accept faster. Confirm that a count is what the NuSEDS column actually stores before keeping it.

The statistical modifier is the one kind of "how" that stays in the dictionary, because a *mean* weight and a *maximum* weight are genuinely different variables:

```output
── escapement · NATURAL_SPAWNERS_TOTAL · statistical_modifier ────────────────
   field:   column_dictionary.csv · statistical_modifier_iri
   current: REVIEW: https://w3id.org/smn/TotalStatisticalModifier

   [1] Total   smn   score 15.15
       The reported value is the sum or cumulative total of the summarized
       observations, such as an annual total catch or cumulative escapement.
       https://w3id.org/smn/TotalStatisticalModifier
       review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "statistical_modifier", rank = 1)
```

The candidate is offered because the column name contains *TOTAL*. That is a name, not a fact about the values. Confirm with the data holder that the column really is a sum before accepting it; if you cannot confirm it, reject the slot rather than inheriting a claim from a column name.

### The table slot has no column

One queue entry is not about a column at all. `tables.csv$observation_unit_iri` records what one row of the table represents, so the printed call names the table instead:

```output
── escapement · entity ───────────────────────────────────────────────────────
   field:   tables.csv · observation_unit_iri
   current: REVIEW: https://w3id.org/smn/Escapement

   [1] Escapement   smn   score 14.4
       The number of mature salmon that pass through (or escape) fisheries
       and return to fresh water to spawn.
       https://w3id.org/smn/Escapement
       review <- accept_suggestion(review, role = "entity", rank = 1, table = "escapement")
```

Paste it as printed. `columns =` filters on column names, so a `columns` filter always hides this slot; build the queue without one when you want to reach it.

### Methods do not live in the column dictionary

A method describes how a value was produced, not what the column *is*, so it is recorded where it is actually constant: a procedure shared by a whole table goes in `tables.csv$method_iri`; a citable protocol document goes in `protocol_iri`/`protocol_citation` on `tables.csv` or `dataset.csv`; and a method that varies row by row becomes a code column whose values resolve through `codes.csv$term_iri` to shared-vocabulary procedures.

In the Fraser Coho package, `ESTIMATE_METHOD` is exactly that row-varying case, and metasalmon's deterministic NuSEDS crosswalk has already filled several of its code rows during package creation — `Area Under the Curve` resolves to `https://w3id.org/gcdfo/salmon#AreaUnderTheCurve` with no `REVIEW:` prefix at all. Those rows are written directly and produce no suggestions, so **they never appear in the review queue**. Open `metadata/codes.csv` and check them by eye. A crosswalk is a defensible default, not a decision you made.

Important boundary: `property_iri`, `entity_iri`, `constraint_iri`, and `statistical_modifier_iri` are I-ADOPT-style measurement-variable components. `unit_iri` is measurement semantics too, but units are not an I-ADOPT role, and methods are not part of the variable at all. Do not use these fields as general-purpose relationship fields for identifiers, attributes, or categorical columns.

## Write the decisions back

Decisions live only in the review object until you apply them.

::::::::::::::::::::::::::::::::::::: group-tab

### R

```{r, eval = FALSE, purl = FALSE}
apply_sdp_semantics(pkg_path, review)
```

```output
✔ Applied 8 semantic review decisions to 'output/fraser-coho-example-sdp'.
ℹ Check the result with `validate_salmon_datapackage(path, require_iris =
  TRUE)`.
```

### Python

Pending. Edit the target metadata CSV fields directly, then revalidate.

### Spreadsheet

Save the metadata CSVs you edited, keeping the header row and column order unchanged.

::::::::::::::::::::::::::::::::::::::::::::::::

`apply_sdp_semantics()` is deliberately narrow, and the narrowness is the safety property:

- accepted IRIs are written with the `REVIEW:` prefix stripped;
- rejected slots are cleared;
- **undecided slots are left exactly as they were**, `REVIEW:` markers and all;
- `term_type` is kept in step with `term_iri`, so the dictionary never claims a kind for a term it no longer holds;
- `datapackage.json` and `semantic_suggestions.csv` are updated in the same transaction, because the descriptor duplicates the dictionary's IRI fields and a half-applied edit would leave the package quietly self-inconsistent; and
- **no data CSV byte is touched.**

That last point is the difference from `write_salmon_datapackage()`, which rebuilds a whole package from in-memory objects. Applying a review is a surgical metadata edit. Applying the same review twice produces identical bytes, which is what makes it safe to leave in a build script that runs every day.

Check the claim rather than believing it — the file hash of `data/escapement.csv` is the same before and after:

```{r, eval = FALSE, purl = FALSE}
before <- tools::md5sum(file.path(pkg_path, "data", "escapement.csv"))

apply_sdp_semantics(pkg_path, review)

identical(before, tools::md5sum(file.path(pkg_path, "data", "escapement.csv")))
#> TRUE
```

Rebuild the queue afterwards and it is shorter: the slots you decided now hold final IRIs, so they drop out.

```{r, eval = FALSE, purl = FALSE}
review_semantics(pkg_path)
```

## The finished script

Session 3 said every accepted decision belongs in the script rather than in a manual edit inside `output/`. This is what that looks like for semantics. Append it to `scripts/build_sdp.R`, below the `create_sdp()` call:

```{r, eval = FALSE, purl = FALSE}
review <- review_semantics(pkg_path)

review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "variable", rank = 1)
review <- accept_suggestion(
  review, "NATURAL_SPAWNERS_TOTAL", "property",
  iri = "https://w3id.org/smn/Abundance"          # the shortlist offered the whole variable
)
review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "entity", rank = 1)
review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "unit", rank = 1)
review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "constraint", rank = 1)
review <- accept_suggestion(review, "NATURAL_SPAWNERS_TOTAL", "statistical_modifier", rank = 1)
review <- accept_suggestion(review, role = "entity", rank = 1, table = "escapement")

review <- reject_suggestion(
  review, "AREA", "variable",
  reason = "DFO statistical area, not a body of water"
)

apply_sdp_semantics(pkg_path, review)

validate_salmon_datapackage(pkg_path, require_iris = FALSE)
```

Every line is a decision, every decision can carry its reasoning in a comment beside it, and the whole thing reruns. That is the property the spreadsheet could not give you.

Expect `validate_salmon_datapackage()` to still warn here: the slots you did not decide keep their `REVIEW:` markers, and the free-text placeholders are untouched. That is the correct state for a package mid-review.

One caveat about rerunning. The decisions replay deterministically, but the *candidates* do not: shared vocabularies evolve, so a future `create_sdp(seed_semantics = TRUE)` may retrieve a different shortlist and `rank = 1` may then name a different term. Where an accepted IRI must never drift, write it as `iri = "…"` instead of `rank = 1`. That converts a position into a commitment.

## What this review cannot decide

Say these out loud in the workshop. A tool that hides its limits teaches learners to trust it in the places it is weakest.

**A slot with no candidate never appears.** The queue is built from shortlists, not from gaps. A required field that retrieval found nothing for is silently absent from the review, so a completed review can still fail `require_iris = TRUE`. The validation report — not the empty queue — is the authority on whether the package is finished.

**An empty queue is not proof of completion.** Filter to a column that does not exist and the console still reports "No unfilled semantic slots with suggestions", which reads like success. Check the column name before believing it.

**Free-text fields are still edited by hand.** Descriptions, contacts, licences, and the other `MISSING …:` placeholders are outside this flow; strict validation reports them as errors. Session 6 covers that gate.

**Rejections do not persist into a fresh queue.** `apply_sdp_semantics()` records the rejection in `semantic_suggestions.csv` and clears the field — but a blank field is undecided, so calling `review_semantics()` again re-queues the slot you rejected, with no sign that you already looked at it. Your script is the record; the queue is not. This is a real rough edge, and it is another reason the decisions belong in `scripts/build_sdp.R`.

**Some suggestion targets are not decidable here at all.** A target that has no single IRI field to write into — the dataset keyword list, for example — is reported and skipped rather than shown as something you can accept. Edit those in the metadata CSVs directly.

## Optional LLM-assisted bundle review

This is a **separate step, and it is strictly opt-in.** Nothing in this chapter contacts an LLM: `review_semantics()`, `accept_suggestion()`, `reject_suggestion()`, and `apply_sdp_semantics()` make no network call of any kind. LLM assessment happens earlier, during `suggest_semantics()` or `create_sdp()`, and only when you pass `llm_assess = TRUE` in R or `llm_assess=True` in Python together with an approved provider and model. Supplying context files alone never triggers a call.

When suggestions were generated that way, the review console surfaces the assessment under the candidate as an extra `llm:` line with its decision, confidence, and rationale. It surfaces it; it never generates it. If you did not opt in, those lines simply are not there.

What the assessment is allowed to do is judge the retrieved candidates. It cannot invent an IRI. A measurement's variable, property, entity, unit, constraint, and statistical-modifier candidates are reviewed together as one bundle, and a statistical-modifier candidate is proposed only when the column's name, label, or description names an aggregation. Deterministic validators can downgrade an unsupported `accept` to `review`, but they never substitute a term.

Context files must be passed as existing local file paths through `llm_context_files`, never as parsed data frames or documents. Use only an approved provider and non-sensitive context. An LLM decision is evidence for your review, not a replacement for it: you still paste the call.

## Function reference

Full argument lists and return values: [`review_semantics()`][metasalmon-review-semantics], [`accept_suggestion()` and `reject_suggestion()`][metasalmon-accept-suggestion], [`apply_sdp_semantics()`][metasalmon-apply-sdp-semantics].

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 1: Complete the bounded Fraser Coho review

Work with the shared NuSEDS Fraser Coho package, even if you brought your own dataset.

1. Build the queue with `review_semantics()` and count the slots.
2. Decide the six `NATURAL_SPAWNERS_TOTAL` slots: variable, property, entity, unit, constraint, statistical modifier. Use `iri =` where the shortlist is wrong, and write one sentence in the script saying why.
3. Decide the table observation-unit slot.
4. Reject `AREA` with a reason.
5. Apply the decisions and confirm `data/escapement.csv` is byte-identical.
6. Open `metadata/codes.csv` and check the `ESTIMATE_METHOD` rows the crosswalk filled. Which would you keep? Which would you question?

Do not map additional fields during this exercise.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

## Challenge 2: Prove the review is reproducible

Delete `output/fraser-coho-example-sdp/` and rerun `scripts/build_sdp.R` from the top. Budget a few minutes: this reruns the seeding search as well as the review.

- Does the package come back with the same accepted IRIs?
- Which of your decisions would survive a vocabulary change, and which are pinned to `rank = 1`?
- Rerun `apply_sdp_semantics()` a second time without changing anything. What changed on disk?
- Rebuild the queue after applying. Which slots came back, and why?

Then answer the question this chapter exists for: **if you left this project today, what would the next person be able to reconstruct — and what would they have to guess?**

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- The console prints the exact `accept_suggestion()` call; pasting it into your build script is what makes the decision reproducible.
- `review_semantics()` reads suggestions that already exist. It never searches, and it never contacts an LLM.
- `REVIEW:` and a blank field both mean undecided; a retrieval score is not agreement.
- The workshop review is bounded to one measurement, its I-ADOPT decomposition, its unit and statistical modifier, one table observation unit, and one deliberate rejection.
- I-ADOPT roles describe the measurement variable; units are separate, and methods live at the table, protocol, or code level rather than in the dictionary.
- `apply_sdp_semantics()` is a surgical, re-runnable metadata edit that leaves the data files byte-identical.
- The queue shows shortlists, not gaps — the validation report decides whether the package is finished.
- Optional LLM assessment is generated earlier and only on explicit opt-in; the review flow only displays it.

::::::::::::::::::::::::::::::::::::::::::::::::
